---
title: "DANA 4840 Project - Research Question 2"
author: "Aryan Mukherjee, Maryam Gadimova, Patricia Tating, Roman Shrestha"
output: html_document
---

How can we identify and interpret similarities in features within clustering results to enhance the understanding of cluster formation and structure?

## Loading the Libraries
```{r, message=FALSE, warning=FALSE}
library("tidyverse")
library("factoextra")
library("dendextend")
library("cluster")
```

## Reading the Data
```{r}
mtcars <- read.csv("data/mtcars.csv", header = T, sep = ",")
head(mtcars)
```

## Checking Data Structure
```{r}
dim(mtcars)
str(mtcars)
```
We can see that our data comprises 32 observations of different car models and 12 automobile features of mixed (numeric, integer, character) data types.


## Data Pre-processing
```{r}
mtcars_categorical <- data.frame(
  cyl = mtcars$cyl,
  vs = mtcars$vs,
  am = mtcars$am,
  gear = mtcars$gear,
  carb = mtcars$carb
)

mtcars_numerical <- data.frame(
  mpg = mtcars$mpg,
  disp = mtcars$disp,
  hp = mtcars$hp,
  drat = mtcars$drat,
  wt = mtcars$wt,
  qsec = mtcars$qsec
)

mtcars_numerical_scaled <- data.frame(scale(mtcars_numerical))

mtcars_joined <- cbind(mtcars_numerical_scaled, mtcars_categorical)
rownames(mtcars_joined) <- mtcars$model
mtcars <- mtcars_joined
head(mtcars)
```

### Average Linkage Method
```{r}
hc_average <- hclust(d = res.dist, method = "average")
grp_average <- cutree(hc_average, k = 2)

average_cluster <- fviz_cluster(
  list(data = mtcars, cluster = grp_average),
  palette = "jco",
  geom = "point",
  ellipse.type = "convex",
  show.clust.cent = FALSE,
  main = "Average Linkage Cluster",
  ggtheme = theme_minimal()
)

average_dendrogram <- fviz_dend(
  hc_average,
  cex = 0.5,
  k = 2,
  k_colors = "jco",
  rect = TRUE,
  rect_border = "jco",
  rect_fill = TRUE,
  label_cols = "black",
  label_cex = 0.5,
  main = "Average Linkage Dendrogram",
  ggtheme = theme_minimal()
)
average_dendrogram
```

## Clustering Results
```{r}
grp <- cutree(hc_average, k = 2)
head(grp, n = 32)
```

```{r}
table(grp)
```

```{r}
fviz_cluster(list(data = mtcars, cluster = grp),
             palette = c("#2E9FDF", "#E7B800"),
             ellipse.type = "convex", #Concentration ellipse
             repel = TRUE, #Avoid label overplotting (slow)
             show.clust.cent = FALSE, ggtheme = theme_minimal()
)
```



```{r}
mtcars_num_results <- cbind(
  mtcars_numerical,
  hclust_average = grp
)
```

```{r}
par(mfrow = c(2, 3))

mtcars_results_col <- colnames(mtcars_num_results)[-length(colnames(mtcars_num_results))]

for (i in seq_along(mtcars_results_col)) {
  column_name <- mtcars_results_col[i]

  boxplot(
    mtcars_num_results[[column_name]] ~ mtcars_num_results$hclust_average,
    xlab = "H Cluster",
    ylab = column_name,
    main = paste(column_name),
    col = color_palette[i %% length(color_palette) + 1]
  )
}

par(mfrow = c(1, 1))
```
```{r}
mtcars_cat_results <- cbind(
  mtcars_categorical,
  hclust_average = grp
)
mtcars_cat_results
```

```{r}
mtcars_cat_results$hclust_average <- as.factor(mtcars_cat_results$hclust_average)

variables_to_plot <- colnames(mtcars_cat_results)[colnames(mtcars_cat_results) != "hclust_average"]

plot_cyl <- ggplot(mtcars_cat_results, aes(x = factor(cyl), fill = factor(hclust_average))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sample(color_palette, 6)) +
  theme_classic() +
  ggtitle("Cylinders")

plot_vs <- ggplot(mtcars_cat_results, aes(x = factor(vs), fill = factor(hclust_average))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = samplesample(color_palette, 4)) +
  theme_classic() +
  ggtitle("Engine Shape (vs)")

plot_am <- ggplot(mtcars_cat_results, aes(x = factor(am), fill = factor(hclust_average))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sample(color_palette, 4)) +
  theme_classic() +
  ggtitle("Transmission (am)")

plot_gear <- ggplot(mtcars_cat_results, aes(x = factor(gear), fill = factor(hclust_average))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sample(color_palette, 4)) +
  theme_classic() +
  ggtitle("Gears")

plot_carb <- ggplot(mtcars_cat_results, aes(x = factor(carb), fill = factor(hclust_average))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sample(color_palette, 12)) +
  theme_classic() +
  ggtitle("Carburetors")

# Arrange the plots in a grid
grid.arrange(plot_cyl, plot_vs, plot_am, plot_gear, plot_carb, ncol = 2)

```

**Cluster 1:** Cars in this cluster typically have:

- Lower horsepower (hp)
- Higher fuel efficiency (mpg)
- Lower weight (wt)
- Smaller engine displacement (disp)
-Fewer cylinders (cyl)

These characteristics suggest that cars in Cluster 1 are generally more economical and smaller, possibly including compact and subcompact cars.

**Cluster 2:** Cars in this cluster typically have:

- Higher horsepower (hp)
- Lower fuel efficiency (mpg)
- Higher weight (wt)
- Larger engine displacement (disp)
- More cylinders (cyl)

These characteristics suggest that cars in Cluster 2 are generally more powerful and larger, possibly including sports cars, muscle cars, and luxury cars.