---
title: "Untitled"
author: "Roman"
output: html_document
---

## Loading the libraries
```{r}
library("tidyverse")
library("factoextra")
library("dendextend")
library("hopkins")
library("corrplot")
library("cluster")
```


## Loading the Datasets
### K-means Clustering: wdbc
```{r}
wdbc <- read.table("data/wdbc.data", header = T, sep = ",")
wdbc
```

### Hierarchical Clustering: mtcars
```{r}
mtcars <- read.csv("data/mtcars.csv", header = T, sep = ",")
head(mtcars)
```

## Pre-processing and Normalizing the data
```{r}
mtcars_categorical <- data.frame(
  cyl = factor(mtcars$cyl),
  vs = factor(mtcars$vs),
  am = factor(mtcars$am),
  gear = factor(mtcars$gear),
  carb = factor(mtcars$carb)
)

mtcars_numerical <- data.frame(
  mpg = mtcars$mpg,
  disp = mtcars$disp,
  hp = mtcars$hp,
  drat = mtcars$drat,
  wt = mtcars$wt,
  qsec = mtcars$qsec
)

mtcars_numerical_scaled <- data.frame(scale(mtcars_numerical))

cars <- cbind(mtcars_numerical_scaled, mtcars_categorical)
rownames(cars) <- mtcars$model
head(cars)
```


```{r}
str(cars)
```


```{r}
rownames(cars)
```




## 1. Assessing Clustering Tendency
#### 1.1. Statistical methods
```{r}
# Compute Hopkins statistic for cars dataset
set.seed(101)

hopkins_val <- hopkins(mtcars_numerical_scaled, m = ceiling(nrow(mtcars_numerical_scaled) / 10))
hopkins_val
```


#### 1.2. Visual methods
```{r}
fviz_dist(
  dist(cars),
  show_labels = FALSE
) + labs(title = "Cars Data")
```

## 2. Applying Various Linkage methods
#### 2.1. Single Linkage Method
```{r}
res.dist <- dist(cars, method = "euclidean")
res.dist
```


```{r}
as.matrix(res.dist)[1:6, 1:6]
```


```{r}
hc_single <- hclust(d = res.dist, method = "single")
hc_single
```


```{r}
# Set the labels to the car models
# hc_single$labels <- rownames(mtcars)

fviz_dend(
  hc_single,
  cex = 0.5,
  main = "Single Linkage Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_single))
```


```{r}
# Plot the dendrogram with custom labels and increased visibility
fviz_dend(
  hc_single,
  cex = 0.5,                # Increase the label size
  k = 4,                              # Number of clusters
  k_colors = "jco",                   # Color palette for clusters
  rect = TRUE, rect_border = "jco",   # Add rectangle around clusters
  rect_fill = TRUE,                   # Fill the rectangle
  label_cols = "black",               # Color of labels
  label_cex = 0.5,                    # Font size of labels
  main = "Single Linkage Dendrogram"
) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


#### 2.2. Complete Linkage Method
```{r}
hc_complete <- hclust(d = res.dist, method = "complete")
hc_complete
```


```{r}
fviz_dend(
  hc_complete,
  cex = 0.5,
  main = "Complete Linkage Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_complete))
```


#### 2.3. Average Linkage Method
```{r}
hc_average <- hclust(d = res.dist, method = "average")
hc_average
```


```{r}
fviz_dend(
  hc_average,
  cex = 0.5,
  main = "Average Linkage Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_average))
```


#### 2.4. Centroid Linkage Method
```{r}
hc_centroid <- hclust(d = res.dist, method = "ward.D2")
hc_centroid
```


```{r}
fviz_dend(
  hc_centroid,
  cex = 0.5,
  main = "Centroid Linkage Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_centroid))
```


#### 2.5. ward.D Method
```{r}
hc_ward <- hclust(d = res.dist, method = "ward.D")
hc_ward
```


```{r}
fviz_dend(
  hc_ward,
  cex = 0.5,
  main = "Ward Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_ward))
```


#### 2.6. ward.D2 Method
```{r}
hc_ward2 <- hclust(d = res.dist, method = "ward.D2")
hc_ward2
```


```{r}
fviz_dend(
  hc_ward2,
  cex = 0.5,
  main = "Ward Dendrogram"
) + theme(axis.text.x = element_text(angle = 0, hjust = 2))
```


```{r}
# Correlation between cophenetic distance and the original distance
cor(res.dist, cophenetic(hc_ward2))
```


## 3. Comparison between Dendrogram
```{r}
# Create multiple dendrograms by chaining
dend_complete <- cars %>%
  dist %>%
  hclust("complete") %>%
  as.dendrogram
dend_single <- cars %>%
  dist %>%
  hclust("single") %>%
  as.dendrogram
dend_average <- cars %>%
  dist %>%
  hclust("average") %>%
  as.dendrogram
dend_centroid <- cars %>%
  dist %>%
  hclust("centroid") %>%
  as.dendrogram
dend_ward <- cars %>%
  dist %>%
  hclust("ward.D") %>%
  as.dendrogram
dend_ward2 <- cars %>%
  dist %>%
  hclust("ward.D2") %>%
  as.dendrogram
```


```{r}
# Compute Correlation Matrix
dend_list <- dendlist(
  "Complete" = dend_complete,
  "Single" = dend_single,
  "Average" = dend_average,
  "Centroid" = dend_centroid,
  "Ward" = dend_ward,
  "Ward2" = dend_ward2
)

cors <- cor.dendlist(dend_list)

# Print correlation matrix
round(cors, 2)
```


```{r}
# Visualize the correlation matrix using corrplot package
corrplot(cors, "pie", "lower")
```


```{r}
# Agglomerative Nesting (Hierarchical Clustering)
res.agnes <- agnes(
  x = cars, # data matrix
  stand = TRUE, # Standardize the data
  metric = "euclidean", # metric for distance matrix
  method = "ward" # Linkage method
)
```


```{r}
# Divisive Analysis Clustering
res.diana <- diana(
  x = cars, # data matrix
  stand = TRUE, # standardize the data
  metric = "euclidean" # metric for distance matrix
)
```


```{r}
fviz_dend(res.agnes, cex = 0.6, k = 4)
```


```{r}
fviz_dend(res.diana, cex = 0.6, k = 4)
```


```{r}
fviz_dend(
  res.hc,
  k = 5, # Cut in four groups
  cex = 0.5, # label size
  palette = "jco",
  color_labels_by_k = TRUE, # color labels by groups
  rect = TRUE # Add rectangle around groups
)
```

